---
title: "MCD607 - Project 1 - Chess"
author: "Marco Castro"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(gt)
```

## Overview

For this project, we were given a text document with a very specific formatting and we were asked to reformat into dataframe of columns "Id", "Player", "State", "Total", "Player Pre-Tournament Rtg", "Avg. Openent Tournament Rating" that can exported as a csv.

I read the text file in two ways: 

- using _readLines_, 
- using _read.delim_

First, I initialized a variable for my filename, a list of column headers, and number of columns to use later on.

```{r init-variables}

file <-'tournamentinfo.txt'

headers <- c("player_id", "Player", "Total", "Round 1", "Round 2", "Round 3", "Round 4", "Round 5", "Round 6", "Round 7", "State", "USCFID/Rtg (Pre->Post)", "n", "Side 1", "Side 2", "Side 3", "Side 4", "Side 5", "Side 6", "Side 7")

num_cols <- length(headers)
```

## Reading Text File Using as _readLines_

The _readLines_ function provides a simple method for reading the file as type char. After reading in the file, I began to parse the data. First, I used grep to remove any line that includes the record breaks ("---"). I then split each line into columns using the _strsplit_ function and setting my separator to "|". This resulted in a list. 

```{r read-txt-lines}
# readlines 
raw_lines <- readLines(file)

# subset only data rows by omitting the breaker lines
data_rows <- raw_lines[grep("---", raw_lines, invert = TRUE)]
# split by |
rows_split = strsplit(data_rows, "\\|")
typeof(rows_split)
```

### Joining data from two rows

The resulting list still had each player's information spread accross two rows. To stitch the rows into a single record, I iterated through the list with a for loop to create a reformated list. Since the for loop checks every row, I used the modulo match function to check for odd and even rows. When the loop hit an odd row, I merged it with the next row (an even row) and added this new list to another list called _parsed_list_.

```{r parse-line_data}
# Loop through all rows to rebuild list,
# joining sets of odd and even rows; 
# start with 3 to skip headers
parsed_list  <- list()
for (i in 3:length(data_rows)) {
  
  # on odd rows, combine current and next row and add to new list
  if(i %% 2 == 1) {
    odd_row <- unlist(rows_split[i])
    evn_row <- unlist(rows_split[i + 1])
    row_list <- append(odd_row, evn_row)
    parsed_list <- c(list(row_list), parsed_list)
  }
}
```
### Converting List to DataFrame
Once I had the data per player onto a single line, I needed to convert it to dataframe. I first used _unlist_ to expand my data into a long format, putting every value on a single line. I then used the _matrix_ function into the right shape with 20 cols by 64 rows. I casted the result into a dataframe and set my headers.

```{r convert-list-to-df}
# convert list to dataframe
lines_df <- data.frame(matrix(unlist(parsed_list), nrow=length(parsed_list), ncol=num_cols, byrow = 1))
# add column names
names(lines_df) <- headers

# convert dataframe to tibble and trim white space
line_data <- as_tibble(lines_df) |>
   mutate(across(where(is.character), str_trim))
head(line_data)

```

## Reading using _read.delim_

Although my data ended up in the correct format, I wasn't sure if I loved the approach. I had read that there were performance benefits of using dplyr and tidyr over for loops. I opted to give it another shot, this time using the _read.delim_ function declare my separators and remove blank columns ahead of time. This method also had the added benefit of returning a dataframe, so I didn't need to convert it as in the last step. The record breaks ("---") converted in NAs so I used _drop_na_ to remove these rows from my dataset.

```{r read_delim_ex}
# read text file usig read.delim
delim_data <- read.delim(file, header=FALSE, sep="|", skip=4, skipNul=1, blank.lines.skip=1, na.strings=c(""," ","NA")) 

# drop all NA's
delim_data_df <- Filter(function(x)!all(is.na(x)), delim_data) |>
  drop_na()
```
### Consolidating Rows
Similar to my for loop test, I used the modulo within the _slice_ function to split my data rows into two dataframes, each containing half of the record's data. I added a column for the row number, which I used as my key when stitching the dataframes back together using _merge_ function. I later dropped the row_id column and declared my new dataframes column names. There's probabbly a more elegant way to do this, but I wasn't able to find a method that worked.

```{r merging-rows-to-create-records}
# split data into two dfs
df_even <- delim_data_df |>
  slice(which(row_number() %% 2 == 1)) |>
  mutate(row_id = row_number())

df_odd <- delim_data_df |>
  slice(which(row_number() %% 2 == 0)) |>
  mutate(row_id = row_number())

# merge data by row_id
chess_data <- merge(df_even, df_odd, by="row_id") |>
  subset(select = -c(row_id)) 

# add column names
names(chess_data) <- headers
head(chess_data)
```

## Tidying Up the Dataframe
I next had to begin tidying up the dataframe. First, I needed to extract the multiple values stored in the "USCFID/Rtg (Pre->Post)" column, including the pre-tournament rating column. I used the _separate_ function in several steps to break up "USCFID/Rtg (Pre->Post)" to "USCFID", 'Rtg (Pre)', 'Rtg (Post)',  'Rtg (Pre P)',  'Rtg (Post P)'. R displayed some warnings since the "P" column had many missing values, but these can be ignored since we won't be using these columns. I created a subset with only the values I needed and assigned this to the dataframe 'chess_player_rating'.

I also needed to do something similar with the Round # columns, as this column stored both the outcome (W=Win, L=Loss, H=Half Point, U=Withdrew).
```{r tidy-up}

# Break up "USCFID/Rtg (Pre->Post)" into its parts
chess_player <- chess_data |>
  separate(col="USCFID/Rtg (Pre->Post)", into=c('USCFID', 'Rtg'), sep = "\\/ R:") 
  
chess_player_rating <- chess_player |>
  separate(col="Rtg", into=c('Rtg (Pre)', 'Rtg (Post)'), sep = "->") |>
  separate(col="Rtg (Pre)", into=c('rtg_pre', 'Rtg (Pre P)'), sep = "P") |>
  separate(col="Rtg (Post)", into=c('rtg_post', 'Rtg (Post P)'), sep = "P") |>
  mutate(across(where(is.character), str_trim)) |>
  mutate(
    rtg_pre = as.integer(rtg_pre),
    rtg_post = as.integer(rtg_post)
  ) |>
  select("player_id", "Player", "State", "Total", rtg_pre, rtg_post)
```

I also needed to do something similar with the Round # columns, as this column stored both the outcome (W=Win, L=Loss, H=Half Point, U=Withdrew) and the opponent's id. I used _pivot_longer_ to get every observation into its own row, then separate to break up the result and opponent's id into their own colummn. I removed any white space then filtered for W|L rows as H|U represent unplayed matches. I did a right_join to my player ratings dataframe using the oppenent_id as a key. Finally, I ran a summary to get mean or average oppenent rating pre-tournament.


```{r tidy-up-2}
# get opponent pre ratings per match
tournament_opponents_ratings <- chess_player |>
  select( "player_id","Round 1", "Round 2", "Round 3", "Round 4", "Round 5", "Round 6", "Round 7") |>
  pivot_longer(
    cols = !(player_id),
    names_to = c( "round"), 
    values_to = "results"
  ) |>
  separate(col="results", into=c("result","opponent_id"), sep="(\\s{2})") |>
  mutate(across(where(is.character), str_trim)) |>
  filter(result == "W" | result =="L") |>
  right_join(chess_player_rating, join_by(opponent_id == player_id))

# summarise oppent ratings per player
opponent_pre_rtgs <- tournament_opponents_ratings  |>
  group_by(player_id) |>
  summarise(
    avg_opponent_rtg =  mean(rtg_pre),
    avg_opponent_rtg = round(avg_opponent_rtg,0)
  )

```
## Merging the results
Having my player ratings and their openent ratings, I conducted a _left_merge_ using the player_id as my primary key to combine the two datasets into their final form.


```{r join-sets}
# merge final dataframe
chess_player_stats <- chess_player_rating |>
  left_join(opponent_pre_rtgs, join_by(player_id == player_id)) |>
  select("Player", "State", "Total", rtg_pre, avg_opponent_rtg)

gt(chess_player_stats)

```
## Writing the Dataframe to CSV
Finally, I used write.csv to export my dataframe as a csv.

```{r write-csv}
write.csv(chess_player_stats,"chess_tournament_player_stats.csv")
```
