---
title: "D607 - Assignment 3 - Normalization and Reg Exp"
author: "Marco Castro"
date: "2024-09-21"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(scales)
```


# Overview
For assignment #3, we focused on normalization and regular expressions.

## 1. Normalization
This section focuses on normalizing dataframes. First I constucted the three dummy dataframes focusing on subset of Major League baseball teams, their respective stadiums and leagues, and the last 4 games (a series) played using the tribble functions.

```{r normalized-dataframes}

# original dataset
games <- tribble(
  ~game_id, ~date, ~visitor, ~home, ~visitor_score, ~home_score,
  1, "2024-09-18", "Pirates", "Reds", 3, 4,
  2, "2024-09-18", "Nationals", "Cubs", 7, 2,
  3, "2024-09-18", "Tigers", "Orioles", 1, 0,
  4, "2024-09-18", "Blue Jays", "Rays", 2, 3,
  5, "2024-09-18", "Phillies", "Mets", 4, 3,
  6, "2024-09-19", "Pirates", "Reds", 1, 2,
  7, "2024-09-19", "Nationals", "Cubs", 2, 0,
  8, "2024-09-19", "Tigers", "Orioles", 2, 3,
  9, "2024-09-19", "Blue Jays", "Rays", 1, 5,
  10, "2024-09-19", "Phillies", "Mets", 8, 10,
  11, "2024-09-20", "Pirates", "Reds", 2, 3,
  12, "2024-09-20", "Nationals", "Cubs", 2, 6,
  13, "2024-09-20", "Tigers", "Orioles", 4, 1,
  14, "2024-09-20", "Blue Jays", "Rays", 5, 3,
  15, "2024-09-20", "Phillies", "Mets", 2, 12,
  16, "2024-09-21", "Pirates", "Reds", 1, 7,
  17, "2024-09-21", "Tigers", "Cubs", 5, 1,
  18, "2024-09-21", "Tigers", "Orioles", 6, 4,
  19, "2024-09-21", "Blue Jays", "Rays", 2, 3,
  20

# create division tibble
division <- tribble(
  ~division_id, ~league, ~division, 
  "ALE", "American", "East",
  "ALC", "American", "Central",
  "ALW", "American", "West",
  "NLE", "National", "East",
  "NLC", "National", "Central",
  "NLW", "National", "West"
)

# create teams tibble
teams <- tribble(
  ~team_id, ~name, ~division,
  "PHI", "Phillies", "NLE",
  "NYM", "Mets", "NLE",
  "TOR", "Blue Jays", "ALE",
  "TB", "Rays", "ALE",
  "BAL", "Orioles", "ALE",
  "DET", "Tigers", "ALC",
  "WSH", "Nationals", "NLE",
  "CHC", "Cubs", "NLC",
  "CIN", "Reds",  "NLC",
  "PIT", "Pirates",  "NLC"
)

# create stadium tibble
stadium <- tribble(
  ~team_id, ~stadium, ~address, ~city, ~state,
  "PHI", "Citizens Bank Park","1 Citizens Bank Way","Philadelphia","PA",
  "NYM", "Citi Field", "41 Seaver Wy", "Flushing", "NY",
  "TOR", "Rogers Center","1 Blue Jays Way", "Toronto", "ON",
  "TB", "Tropicana Field","1 Tropicana Dr.","Tampa Bay", "FL",
  "BAL", "Camden Yards","333 W Camden St", "Baltimore", "MD",
  "DET", "Comerica Park","2100 Woodward Ave", "Detroit", "MI",
  "WSH", "Nationals Park", "1500 South Capitol St SE", "Washington", "DC",
  "CHC", "Wrigley Field","1060 W Addison St", "Chicago", "IL",
  "CIN", "Great American Ball Park", "100 Joe Nuxhall Wy", "Cincinnati", "OH",
  "PIT", "PNC Park", "115 Federal St", "Pittsburgh", "PA"
)

# create games tibble
games <- tribble(
  ~game_id, ~date, ~visitor, ~home, ~visitor_score, ~home_score,
  1, "2024-09-18", "PIT", "CIN", 3, 4,
  2, "2024-09-18", "WSH", "CHC", 7, 2,
  3, "2024-09-18", "DET", "BAL", 1, 0,
  4, "2024-09-18", "TOR", "TB", 2, 3,
  5, "2024-09-18", "PHI", "NYM", 4, 3,
  6, "2024-09-19", "PIT", "CIN", 1, 2,
  7, "2024-09-19", "WSH", "CHC", 2, 0,
  8, "2024-09-19", "DET", "BAL", 2, 3,
  9, "2024-09-19", "TOR", "TB", 1, 5,
  10, "2024-09-19", "PHI", "NYM", 8, 10,
  11, "2024-09-20", "PIT", "CIN", 2, 3,
  12, "2024-09-20", "WSH", "CHC", 2, 6,
  13, "2024-09-20", "DET", "BAL", 4, 1,
  14, "2024-09-20", "TOR", "TB", 5, 3,
  15, "2024-09-20", "PHI", "NYM", 2, 12,
  16, "2024-09-21", "PIT", "CIN", 1, 7,
  17, "2024-09-21", "WSH", "CHC", 5, 1,
  18, "2024-09-21", "DET", "BAL", 6, 4,
  19, "2024-09-21", "TOR", "TB", 2, 3,
  20, "2024-09-21", "PHI", "NYM", 3, 6,
)
```
Next I joined the dataframes using left joins. Repeating columns such as team "name" were renamed for clarity of wether a team was the home team hosting the game at their stadium or a visitor.

```{r joining-dfs}
games_in_series <- games |>
  left_join(teams, join_by(home == team_id)) |>
  left_join(teams, join_by(visitor == team_id)) |>
  left_join(stadium, join_by(home == team_id)) |>
  left_join(division, join_by(division.x == division_id)) |>
  left_join(division, join_by(division.y == division_id)) |>
  select(date, home_score, visitor_score, name.x, league.x, division.x, name.y, league.y, division.y, stadium, address, city, state) |>
  rename(
    home_team = name.x, 
    home_league = league.x, 
    home_division = division.x,
    visitor_team = name.y,
    visitor_league = league.y, 
    visitor_division = division.y,
    played_at = stadium
  )
```

Next, I wanted to 

```{r runs-scored}
runs_scored_in_series <- games_in_series |>
  mutate(teams = paste(visitor_team, home_team, sep=" at ")) |>
  group_by(teams) |>
  summarise(
    runs_home_team = sum(home_score),
    runs_visitor_team = sum(visitor_score)
  ) 

runs_scored_in_series

runs_scored_in_series |>
  ggplot(aes(x=teams)) +
    coord_flip() +
    geom_bar(stat='identity', aes(y=runs_home_team, fill=
                     "blue")) +
    geom_bar(stat='identity', aes(y=runs_visitor_team, width = 0.3, fill=
                     "red")) +
    labs(x="Series", y = 'Runs Scored') +
    scale_fill_manual(values = c("blue","red"),labels = c("home","visitor"))


```



## Regular expression 
This section focuses on using regular expressions.

### 2. Character Manipulation

We were asked to download a dataset showing 175 college majors from Five Thirty Eight (<https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/>) and identify the majors  containing "DATA" or "STATISTICS" in their names.

```{r import-prep-data}

# load data from 538 github page
college_majors <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv")

# filter df where majors contain "DATA" OR "STATISTICS" using str_detect
college_majors_data_statistics <- college_majors |>
  filter(str_detect(Major, "DATA|STATISTICS"))

# filter df where majors contain "DATA" OR "STATISTICS" using GREPL
college_majors_data_statistics_grepl <- college_majors |>
  filter(grepl("DATA|STATISTICS", Major))
```

Using the function _str_detect_ within a filter, I was able to determine that the majors with those keywords are:
`r college_majors_data_statistics`

For comparison, I also attempted to filter using grepl and achived the same results
`r college_majors_data_statistics_grepl`

### 3. Description of expressions:
We were asked to define the following regular expressions:

```{r regexp}
p1 <- "(.)\1"
p2 <- "(.)(.)\\2\\1"
p3 <- "(..)\1"
p4 <- "(.).\\1.\\1"
p5 <- "(.)(.)(.).*\\3\\2\\1"
```

_a: (.)\1\1_
Returns true if string contains any characters followed by \1. For example "T\1" or "e\1" in the code block below:

```{r regexp-q1a}
test_str <- c('TTT')
str_detect(test_str, p1)
```

_b: "(.)(.)\\2\\1"_
Returns true if string two characters followed by the same two characters reordered from last to first. For example: "xssx" or "anna" in the code block below:

```{r regexp-q1b}
test_str <- c("TsxssxT", 'apple\1', 'carrot', '\1brocoli', 'bananna')
str_detect(test_str, p2)
```

_c: (..)\1_
Returns true if string contains two characters followed by \1. For example "Ts\1" or "le1" in the code block below:

```{r regexp-q1c}
test_str <- c('TsTs\1sxded', 'apple\1', 'carrot', '\1brocoli', 'banana')
str_detect(test_str, p3)
```

_d:  "(.).\\1.\\1"_
Returns true if string contains a character followed by any character, repeats the first character followed by any character, then repeats the first character again. For example: "AbAcA" in the code block below:

```{r regexp-q1d}
test_str <- c( 'AbAcA', 'apple\1', 'AbAcA', '\1brocoli', 'banana')
str_detect(test_str, p4)
```

_e: "(.)(.)(.).*\\3\\2\\1"_
Returns true if string contains three characters (ABC in my example), followed by 0 or more characters, then repeats the first three characters in the matching expression in reverse order. For example: "ABCsomethingCBA" in the code block below or "ABCsCBA, or "ABCCBA."

```{r regexp-q1e}
test_str <- c('ABCsomethingCBA', 'apple\1', 'carrot', '\1brocoli', 'banana')
str_detect(test_str, p5)
``` 

### 4. Constructing regular expressions

We were asked to construct regular expressions for the following conditions:

_a: Start and end with the same character_
(.).+\\1

```{r regexp-same-char}
str_detect("xHellox", "(.).+\\1")
```

_b: Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)_
(..).+\\1

```{r regexp-contains-repeating-letters}
str_detect("church", "(..).+\\1")
```

_c: Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)_

(.).\\1.\\1

```{r regexp-contains-repeated-letters}
str_detect(c("eleven","bananas"), "(.).\\1.\\1")
```

## Conclusion
Regular expressions are a powerful tool that will come in handy.
