---
title: "Week 10 Assignment"
author: "Marco Castro"
date: "2024-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
library(tidytext)

library(janeaustenr)
```
## Sentiment Analysis
The assignment for Week 10 is broken up into two parts. The first is to replicate the primary example from _Text Mining with R_, (Chapter 2)[https://www.tidytextmining.com/sentiment.html]. The second part of the assignment it to use a different corpus and incorporate at least one additional sentiment lexicon.

### Replicating 

The following code chunk was copied over from the book _Text Mining with R_, (Chapter 2)[https://www.tidytextmining.com/sentiment.html for educational practice purposes. Specifically it uses a corpus of "Pride & Prejudice 

```{r load sentiment}
# Source : Text Mining with R, Chapter 2. https://www.tidytextmining.com/sentiment.html

tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",  ignore_case = TRUE)))
  ) %>%
  ungroup() %>%
  unnest_tokens(word, text)

#loughran

pride_prejudice <- tidy_books %>% 
  filter(book == "Pride & Prejudice")

afinn <- pride_prejudice %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(index = linenumber %/% 80) %>% 
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "AFINN")


bing_and_nrc <- bind_rows(
  pride_prejudice %>% 
    inner_join(get_sentiments("bing")) %>%
    mutate(method = "Bing et al."),
  pride_prejudice %>% 
    inner_join(get_sentiments("nrc") %>% 
                 filter(sentiment %in% c("positive", 
                                         "negative"))
    ) %>%
    mutate(method = "NRC")) %>%
  count(method, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

# plot number of words 
bind_rows(afinn, 
          bing_and_nrc) %>%
  bind_rows(loughran) %>%
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")
```




```{r add-loughran}

loughran <- pride_prejudice %>% 
  inner_join(get_sentiments("loughran")) %>% 
  group_by(index = linenumber %/% 80) %>% 
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "LOUGHRAN")

```

```{r plot-diffs}
bind_rows(afinn, 
          bing_and_nrc) %>%
  bind_rows(loughran) %>%
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")
```


